Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2023-02-10 src/tex/ms.tex
Add edits

In full: we propose a method to ``backward" model each GW event to its
progenitor state while allowing the hyperparameters to vary across their full
range of physical uncertainty following a two-stage process. First, we solve a
root finding problem to obtain guesses that are likely to produce the desired
system properties in the joint progenitor--hyperparameter space. We then sample
the posterior in the joint space that is induced by the DCO merger observations
using a Markov chain Monte Carlo (MCMC) algorithm that is initialized by the
roots found in the previous stage.  We demonstrate our algorithm in a realistic
setting by producing a posterior over progenitor properties and \texttt{COSMIC}
hyperparameters implied by the observation of
dependencies:
  src/scripts/figure1_cornerPlot.py:
    - src/data/GWTC3_LVK_posterior.npz
    - src/data/GWTC3_mcmc_processed.hdf5

  src/scripts/figure2_varyAlphaFacc.py:
    - src/data/fig_2_dat.h5

  src/scripts/figure3_sanityCheck.py:
    - src/data/GWTC3_LVK_posterior.npz
    - src/data/GW150914_roots_raw.hdf5
    - src/data/GW150914_roots_processed.hdf5
    - src/data/GWTC3_mcmc_processed.hdf5

  src/scripts/figure4_population.py:
    - src/data/GWTC3_mcmc_processed.hdf5

datasets:
  10.5281/zenodo.6624911:
    destination: src/data
    contents:
      GW150914_roots_processed.hdf5:
      GW150914_roots_raw.hdf5:
      GWTC3_LVK_posterior.npz:
      GWTC3_mcmc_processed.hdf5:
      fig_2_dat.h5:
 GW150914.

@@

Including hyperparameters and random variables %in the sampling
drastically increases the dimensionality of the problem, which can make the
sampling process much more computationally expensive to converge. To speed up
the convergence, we first solve a root-finding problem to find points
$\bm{\Theta}$ that will generate GW parameters $\bm{\theta}$ close to the bulk
of the posterior density of a GW event.  We then use those points to initialize
a set of chains in the MCMC process.

@@

% Validity paragraph \wf{What if we made this a subsection of ``methods'', and
%then just one sentence here about the values of the KL divergence for
%GW150914?} To check the validity of the recovered posterior over evolutionary
%parameters, it is essential to check whether it is consistent with the original
%posterior in the observable space. We reproject the recovered posterior to the
%observable space, and compare the KL divergence between the original posterior
%and the reprojected posterior.

@@

For progenitor parameters $\bm{\theta'}$, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.

For hyperparameters $\bm{\lambda}$, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$.

>>>>

In full: we propose a method to ``backward" model each GW event to its
progenitor state while allowing the hyperparameters to vary across their full
range of physical uncertainty following a two-stage process. First, we solve a
root finding problem to obtain guesses that are likely to produce the desired
system properties in the joint progenitor--hyperparameter space. We then sample
the posterior in the joint space that is induced by the DCO merger observations
using a Markov chain Monte Carlo (MCMC) algorithm that is initialized by the
roots found in the previous stage.  We demonstrate our algorithm in a realistic
setting by producing a posterior over progenitor properties and \texttt{COSMIC}
hyperparameters implied by the observation of GW150914.

@@

Including hyperparameters and random variables %in the sampling
drastically increases the dimensionality of the problem, which can make the
sampling process much more computationally expensive to converge. To speed up
the convergence, we first solve a root-finding problem to find points
$\bm{\Theta}$ that will generate GW parameters $\bm{\theta}$ close to the bulk
of the posterior density of a GW event.  We then use those points to initialize
a set of chains in the MCMC process. \textbf{We use \textsc{emcee} \cite{emcee}
to sample the posterior density in the evolutionary parameter space.}

@@

@@

For progenitor parameters $\bm{\theta'}$, we characterize each guess with five
parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the
orbital period $t_{\rm orb}$, the eccentricity $e$, and the metallicity $Z$ at
ZAMS. \textbf{We use a log-uniform prior in sampling $M_{\rm 1,ZAMS}$, $M_{\rm
2,ZAMS}$, $t_{\rm orb}$, and $Z$, and we use a uniform prior for $e$.
The range follows the range for the optimization range as given by table
\ref{tab:parameters}.} Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does
not intrinsically affect the evolution of the binary. Instead, we compute the
time it takes for the binary to evolve from formation to the merger and add this
time to the lookback time when a GW event is observed and limit the total time
to merger to be less than $13.7\,\rm{Gyr}$. Using the total lookback time, we
can compute the redshift of the binary at ZAMS formation following the Planck
2018 cosmological model \citep{Planck2018}. The main benefit of post-processing
the redshift in this way is that we do not rely on a particular assumption of
star formation rate (SFR) distribution when we solve for the posterior
distribution in redshift. This means we can put the prior on formation redshift
(or in general, metallicity-redshfit distribution) in by reweighting the
posterior distribution according to a particular SFR model, and therefore do not
have to rerun the inference every time we change the SFR model. This is
especially helpful since SFR models are still highly uncertain but strongly
impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.

For hyperparameters $\bm{\lambda}$, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common
envelope efficiency $\alpha$, the fraction of mass accreted onto the accretor
during stable mass transfer $f_{\rm acc}$, the critical mass ratio which
determines whether Roche-overflow mass transfer proceeds stably or produces a
common envelope when the donor star is on the Hertzsprung gap $q_{\rm crit,3}$,
and the root-mean-square of the Maxwellian distribution employed for CO natal
kicks, $\sigma$. \textbf{We use a uniform prior for $\alpha$, $f_{\rm acc}$, and
$q_{\rm crit,3}$, and a log-uniform prior for $\sigma$.}

\textbf{We use the same prior for all the events in this study. Note that our
main goal is to see whether the events are compatible with \textsc{COSMIC} on an
individual event basis, we do not consider any population analysis here, and
therefore we do not account for any selection effect.}
